---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

First, establish some settings regarding where the data is located:
```{r}
url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
zipFile = "activity.zip"
dataDirectory = "data"
dataFile = "activity.csv"
```

Then, perform data caching. Check if the required data is already stored in the 
working directory. If the data has already been extracted into the "data" 
directory as a csv file, then nothing needs to be done at this stage. If the 
data esists only as a zip file in the working directory, then extract the data 
into a csv file. And if the data is not present at all, download and extract 
the data automatically from the source URL. 
```{r}
#===============================================================================
# Checks if the csv data exists
dataFilePath = paste(dataDirectory, dataFile, sep="/")
if(!file.exists(dataFilePath)){
    # Create the data directory if not already present
    if(!file.exists(dataDirectory)){
        dir.create(dataDirectory)
    }
    # Download the zip file from URL if its not already in working directory
    if (!file.exists(zipFile)) {
        download.file(url, destfile = zipFile, mode="wb")
    }
    # unzip the contents of the zip file into data directory
    unzip(zipFile, exdir=dataDirectory)
}
```

#===============================================================================
##                             What is mean total number of steps taken per day?
#===============================================================================



#===============================================================================
## What is the average daily activity pattern?
#===============================================================================



#===============================================================================
## Imputing missing values
#===============================================================================



#===============================================================================
## Are there differences in activity patterns between weekdays and weekends?
#===============================================================================
